Attrition in Massive Open Online Courses: Inequalities, Dropout Reasons, and Psychological Factors

### 2014-10-30 paper_v1.Rnw
final additions and edits before sumission

\begin{abstract}
Research on attrition has a long tradition in the field of education. Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in motivation. A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses revealed gender and geographic inequalities in persistence and performance (Study 1). Time was a critical obstacle for many learners, but satisfaction was only weakly related to persistence. A follow-up study with online learners who disengaged from the course (Study 2; $ N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset. Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%). A low capacity for volitional control was identified as a large contributor to this issue. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

@@

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

\begin{table*}[th]
\label{tab:s1sum}
\caption{Summary statistics of learner demographics and behavior in 14 MOOCs}
\small
\center
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelors' or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

\begin{description}
  \item[RQ1] How do prior learner characteristics (demographics, geographic location, enrollment intentions, and previous online courses) influence learners' likelihood of disengaging from a MOOC?
  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
\end{description}

@@

\begin{description}
  \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
\end{description}

@@

\section{Study 1: Understanding Attrition}

@@

Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course. Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile. A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 14 MOOCs in our dataset. Out of the 65,320 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$. Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.
% The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
% The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
% Information about age, gender, and education level were only available for learners who completed the initial course survey.
% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior

@@

\subsubsection{Predictors of Attrition}

We fit logistic mixed-effects (hierarchical) models to a number of binary persistence and performance metrics to estimate individual differences in prior learner characteristics, including demographic and geographic features. Correlations between predictors were low enough to allow for simple interpretation of coefficients. The 14 MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs. Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model. The coefficients represent changes from a baseline (intercept) population: male learners in North America, with an average age, who had not earned a bachelor's degree and had no prior experience with the topic or taken other MOOCs, not intending to complete the course, and prepared to spend only up to two hours per week on the course.

A number of significant and substantial individual differences emerged, some indicative of potential inequalities: Women were 10 to 15\% points less likely than men to persist with lectures and assessments. Women are also less likely to score a grade above the 60$^\text{th}$ percentile than men. Learners in Africa, Asia, and South America were 20 to 60\% points less likely to persist in video watching and achieved lower grades than learners in North America (N.A.). Learners in Europe exhibited 10 to 30\% points higher persistence with lectures and assessemnts than those in N.A.. While learners in Asia showed lower persistence with videos, they were more likely to attempt more assessments than learners from N.A..

@@

<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model for predicting persistence and achievement milestones (prop. videos watched, prop. assessments completed, and grade percentile). The baseline (intercept) are males in N. America of average age who have not earned a bachelors\', no prior experience in the topic, not intending to complete the course, prepared to spend under 2 hours/week, this being their first MOOC. An estimate of -20 indicates a 20\\% points lower probability of achieving a milestone.'>>=
library(ggplot2)
res = read.csv("pred_coefs.csv")
res$y = factor(res$y, levels=c(">30% videos", ">50% videos", ">80% videos", ">60% pctile grade", ">30% assessments", ">50% assessments", ">80% assessments", ">80% pctile grade"))
res$y = factor(res$y, labels=c("> 30% videos", "> 50% videos", "> 80% videos", "> 60th percentile grade", "> 30% assessments", "> 50% assessments", "> 80% assessments", "> 80th percentile grade"))
res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:S.America")))
ggplot(subset(res, coef2!="Intercept"), 
       aes(coef2, 100*est, ymin=100*(est-1.96*se), ymax=100*(est+1.96*se), color=factor(sign))) + 
  geom_hline(xintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
  theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
  scale_color_manual(values=c('red','grey30','green3')) +
  labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
@

\subsubsection{Satisfaction: Personal Success}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

@@

\subsubsection{Reasons for Disengaging}
% RQ3a: What reasons do learners report for disengaging from a MOOC?
% RQ3b: How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?

Research question $RQ3a$ about learners' reasons for dropping out of a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list. The distribution of proportions across fourteen MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots. Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time. Other reasons for attrition were less prevalent, though still reported by around 10\% to 25\% of respondents. Notably, just under 20\% of respondents in a typical course (the median proportion) stopped participating because they had learned what they wanted to learn.

@@

Learners may disengage from the course when they become too occupied by higher-priority tasks. Enrolling in a MOOC neither guarantees that a learner can or intends to allocate enough time to go through the course. According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gain. A low capacity for volitional control \cite{corno2001volitional} could lead learners misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task. This implies that learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.

\section{Study 2: Understanding Disengagement}

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback. To this end, we developed a predictive model to identify at-risk learners and asked them to complete a survey. One goal of this study was to further investigate research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced. Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time. A variety of factors can influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments, are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. For instance, learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV). We constructed the following hypotheses about the presence of these two factors for learners who report having not enough time:

@@

Another goal of Study 2 was to collect data to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners. To this end, we included a self-report measures for each construct in the survey that was sent to at-risk learners.

@@

The objective was to build an `early warning' prediction model that could flag learners who are likely to have dropped out or to drop out in the future. In order for our predictions to be useful, we need to observe the following three features of the dropout prediction problem.

\textit{Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible. The false positive rate (the fraction of non-dropouts incorrectly predicted as droputs) should be low to avoid sending a survey to learners who do not experience challenges in the courses.

\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short. A shorter prediction lag improves the likelihood of successfully intervening to help the learner get back on track.

\textit{Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course. We thus had to train the prediction model using data from previous courses. The performance of the prediction model should be adequate in courses other than those on which it was trained.

The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model).

To constrain prediction lag, we stopped feeding the model with a learner's activity data once they have been absent for 14 consecutive days. During training, if the learner had dropped out but the model had not detected her within 14 days of her last appearnce, the learner is counted as a miss (false negative). This construction causes the training algorithm to choose from among features that only operate within the prediction lag constraints we define.

The model was fit by logistic regression. Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).

\begin{table}[h!]
\caption{Performance of the dropout prediction model}
\label{tab:drop_pred_rocs}
\small
\center
\begin{tabular}{lccc}
\toprule
 & $train$ & $test_1$ & $test_2$ \\
\midrule
AUC &  0.931 & 0.929 & 0.922  \\
\bottomrule
\end{tabular}
\end{table}

Two of the predictors in the fitted model are highly influential. The first is the number of days since the learner has last been active ($x_{1}$). The likelihood that the learner drops out increases substantially if the learner disengages for 14 days or more. However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed ($x_{2}$) prior to the absence; this is the second most influential feature in the model. Equation 1 is an approximation of the model with only these two predictors. A learner was flagged if their dropout probability exceeded a half:

@@

The almost identical results obtained for $test_1$ and $test_2$ imply good transferability of the model.

Out of the 20,048 enrolled learners in the course, we only considered those 10,510 who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out. The survey was sent out to 6,053 leaners (excluding 21 learners with invalid email addresses or duplicate accounts).

@@

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A low response rate was expected for this subpopulation that was defined by low engagement. And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate). Missing values were multiply imputated by predictive mean matching using responses to all survey questions.\footnote{The $R$ mice package was used to perform multiple imputation and subsequent pooled analyses.} All estimates were pooled across ten imputations, with standard errors adjusted for added variance. The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.

@@

The methodology used in this study can be extended to deliver timely and targeted interventions. The predictive model was successful in identifying a large number of leaners who were facing challenges with taking the course. The majority of them were unsuccessful in their own terms. All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners. Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}).  Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their identity or social belonging.

>>>>

\begin{abstract}
Research on attrition has a long tradition in the field of education. Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation. A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses revealed gender and geographic inequalities in persistence and performance (Study 1). Time was a critical obstacle for many learners, but satisfaction was only weakly related to persistence. A follow-up study with online learners who disengaged from the course (Study 2; $N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset. Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%). A low capacity for volitional control was identified as a large contributor to this issue. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

@@

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses overall. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

\begin{table*}[th]
\label{tab:s1sum}
\caption{Summary statistics of learner demographics and behavior in 17 MOOCs}
\small
\center
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelor's or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

\begin{description}
  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
\end{description}

@@

\begin{description}
  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions, and prior experience) influence learners' likelihood of disengaging from a MOOC?
  \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
\end{description}

@@

\section{Study 1: Attrition \& Satisfaction}

@@

Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course. Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile. A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 17 MOOCs in our dataset. Out of the 63,620 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$. Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.\footnote{The $R$ mice package was used to perform multiple imputation and subsequent pooled analyses.}
% The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
% The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
% Information about age, gender, and education level were only available for learners who completed the initial course survey.
% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior

@@

\subsubsection{Individual Differences in Attrition}

We fit logistic mixed-effects (hierarchical) models to a number of binary persistence and performance metrics to estimate individual differences in prior learner characteristics, including demographic and geographic features. Correlations between predictors were low enough to allow for simple interpretation of coefficients. Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{The required data for 3 of the 17 MOOCs could not be obtained in time for submission due to technical issues.} Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model. The coefficients represent changes from a baseline (intercept) population: male learners in North America, with an average age, who had not earned a bachelor's degree and had no prior experience with the topic or taken other MOOCs, not intending to complete the course, and prepared to spend only up to two hours per week on the course.

A number of significant and substantial individual differences emerged, some indicative of potential inequalities: Women were 10 to 15\% points less likely than men to persist with lectures and assessments. Women are also less likely to score a grade above the 60$^\text{th}$ percentile than men. Learners in Africa, Asia, and South America were 20 to 60\% points less likely to persist in video watching and achieved lower grades than learners in North America. Learners in Europe exhibited 10 to 30\% points higher persistence with lectures and assessemnts than those in North America. While learners in Asia showed lower persistence with videos, they were more likely to attempt more assessments than learners from North America.

@@

<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop. videos watched, prop. assessments completed, and grade percentile). The baseline group were male learners in North America of average age who had not earned a bachelor\'s degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it. An estimate of -20 indicates a 20\\% points lower probability of achieving a milestone.'>>=
library(ggplot2)
res = read.csv("pred_coefs.csv")
res$y = factor(res$y, levels=c(">30% videos", ">50% videos", ">80% videos", ">60% pctile grade", ">30% assessments", ">50% assessments", ">80% assessments", ">80% pctile grade"))
res$y = factor(res$y, labels=c("> 30% videos", "> 50% videos", "> 80% videos", "> 60th percentile grade", "> 30% assessments", "> 50% assessments", "> 80% assessments", "> 80th percentile grade"))
res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:S.America")))
ggplot(subset(res, coef2!="Intercept"), 
       aes(coef2, 100*est, ymin=100*(est-1.96*se), ymax=100*(est+1.96*se), color=factor(sign))) + 
  geom_hline(xintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
  theme_minimal() + theme(legend.position="none",axis.title=element_text(vjust=0)) + 
  scale_color_manual(values=c('red','grey30','green3')) +
  labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
@

\subsubsection{Learner Satisfaction}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

@@

\subsubsection{Reasons for Attrition}
% RQ3a: What reasons do learners report for disengaging from a MOOC?
% RQ3b: How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?

Research question $RQ3a$ about learners' reasons for disengaging from a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.\footnote{This researcher-defined list may not have been collectively exhaustive, which is addressed in Study 2.} The distribution of proportions across 17 MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots. Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time. Other reasons for attrition were less prevalent, though still reported by around 10\% to 25\% of respondents. Notably, just under 20\% of respondents in a typical course (the median proportion) stopped participating because they had learned what they wanted to learn.

@@

Learners may disengage from the course when they become too occupied by higher-priority tasks. Enrolling in a MOOC neither guarantees that a learner can or intends to allocate enough time to go through the course. According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gain. A low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task. This implies that learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.

\section{Study 2: Understanding Attrition}

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback. To this end, we developed a predictive model to identify disengaging learners and asked them to complete a survey. One goal of this study was to further investigate research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced. Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time. A variety of factors can influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments, are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. For instance, learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV). We constructed the following hypotheses about the presence of these two factors for learners who report having not enough time:

@@

Another goal of Study 2 was to collect data to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners. To this end, we included a self-report measure for each construct in the survey that was sent to disengaging learners.

@@

We constructed an `early warning' prediction model to flag learners who had disengaged from the course or were about to disengage. The following three criteria were chosen as benchmarks to evaluate the practical and theoretical value of our prediction model.%   In order for our predictions to be useful, we need to observe the following three features of the dropout prediction problem.

\textit{Prediction accuracy}: Prediction recall (fraction of correctly predicted disengaged learners) should be high to deliver the survey to as many disengaged learners as possible. The false positive rate (fraction of active learners who are incorrectly predicted as disengaged) should be low to avoid sending surveys to learners who do not experience challenges in the course.

\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as disengaged should be short. A shorter prediction lag improves the likelihood of successfully intervening to help learners resume the course.

\textit{Transferability}: Whether or not a learner disengages from the course cannot be determined before the end of the course. We thus trained the prediction model using data from previous courses. The performance of the prediction model should be good in courses other than those on which it was trained.

The training set was composed of learners' interaction data (predictors) and disengagement states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see omitted for blind review%\cite{halawa2014dropout}
, for additional information on the prediction model).

To control the length of the prediction lag, the model was only provided with learners' activity data until they were absent for 14 consecutive days. In the training phase, if the learner disengaged but the model did not detect it within 14 days of their last activity, the case was counted as a false negative. This construction results in a model with features that all fall within the defined prediction lag period. The model was fit by logistic regression with two-fold learner-level and course-level cross-validation. Table \ref{tab:predAuc} provides information on the prediction performance, measured by the area under the ROC curve (AUC).

\begin{table}[h!]
\caption{Performance of the attrition prediction model}
\label{tab:predAuc}
\small
\center
\begin{tabular}{lccc}
\toprule
 & $train$ & $test_1$ & $test_2$ \\
\midrule
AUC &  0.931 & 0.929 & 0.922  \\
\bottomrule
\end{tabular}
\end{table}

Two of the predictors in the fitted model are highly influential. The first is the number of days since the learner has last been active ($x_{1}$). The likelihood that the learner drops out increases substantially if the learner disengages for 14 days or more. However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed ($x_{2}$) prior to the absence; this is the second most influential feature in the model. Equation 1 is an approximation of the model with only these two predictors. The fact that almost identical results were obtained for $test_1$ and $test_2$ suggest good transferability of the model. A learner was flagged if their dropout probability exceeded a half:

@@

Out of the 20,048 enrolled learners in the course, only those 10,510 were considered who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out. The survey was sent out to 6,053 learners (excluding 21 learners with invalid email addresses or duplicate accounts).

@@

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A low response rate was expected for this subpopulation that was defined by low engagement. And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate). Missing values were multiply imputated by predictive mean matching using responses to all survey questions. All estimates were pooled across ten imputations, with standard errors adjusted for added variance. The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelor's or more advanced degree.

@@

The methodology used in this study can be extended to deliver timely and targeted interventions. The predictive model was successful in identifying a large number of learners who were facing challenges with taking the course. The majority of them were unsuccessful in their own terms. All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners. Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}). Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their identity or social belonging.
